---
title: "ActivityMonitoringAnalyis"
author: "Emanuel Sousa"
date: "8 de Junho de 2017"
output: html_document
---

#Loading and preprocessing the data

Let's first load the data:

```{r}
df<-read.csv("activity.csv")

df$newDate <- as.Date(df$date, format="%d/%m/%Y")

```

#Question: What is mean total number of steps taken per day?

Histogram of the total number of steps taken each day

```{r}
#Subset and sum by year
stepsDay<-with(df,tapply(df$steps, df$date, sum))

hist(stepsDay,
     xlab="Number of steps",
     col="red",
     main = "Steps per day",
     ylab = "Days"
     )
```

Mean of the total number of steps taken per day

```{r}
# can only consider non nan values
mean(stepsDay,na.rm=TRUE)
```

Median of the total number of steps taken per day

```{r}
# can only consider non nan values
median(stepsDay,na.rm=TRUE)
```

#Question: What is the average daily activity pattern?

Next a time series plot is presented of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
stepsInterval<-with(df,tapply(df$steps, df$interval, sum,na.rm=TRUE))
plot(stepsInterval,type="l",xlab="Intervals", ylab="Average Steps", main = "Average Steps per interval")
```

Here I identify the 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps
```{r}
as.numeric(names(stepsInterval[which.max(stepsInterval)]))
```        
# Imputing missing values

Total number of missing values in the dataset (i.e. the total number of rows with NAs):

```{r}
sum(is.na(df$steps))
```  

For filling missing data I will define the missing values as the median for the interval. To do so I will define a function that for a given steps value checks if it is "na"" and if it is will look on a table for the corresponding median
```{r}
MedianStepsInterval<-with(df,tapply(df$steps, df$interval, median,na.rm=TRUE))

repfunc <- function(steps, interval ){
    if(is.na(steps))
    {
        med <- MedianStepsInterval[names(MedianStepsInterval) == interval]
    }
    else
    {
        med <- steps
    }
    return(med)
}
```  

Create a new dataframe and replace the "na" step values by the median
```{r}
newdf <- df
newdf$steps<-mapply(repfunc, df$steps, df$interval)
```  

New histogram of the total number of steps taken each day

```{r}
#Subset and sum by year
stepsDayn<-with(df,tapply(df$steps, df$date, sum))
par(mfrow=c(1,2))
hist(stepsDay,
     xlab="Number of steps",
     col="red",
     main = "Steps per day (with Nan)",
     ylab = "Days"
     )
stepsDaywn<-with(df,tapply(newdf$steps, newdf$date, sum))
hist(stepsDaywn,
     xlab="Number of steps",
     col="red",
     main = "Steps per day (without Nan)",
     ylab = "Days"
     )
```

Mean and Median of the total number of steps taken per day

```{r xtable, results="asis"}
library(xtable)
#Steps day with nans replaced
stepsDaywn<-with(newdf,tapply(newdf$steps, newdf$date, sum))
a <- mean(stepsDaywn,na.rm=TRUE)
b <- median(stepsDaywn,na.rm=TRUE)

#Steps day with nans
stepsDayn<-with(df,tapply(df$steps, df$date, sum))
c <- mean(stepsDayn,na.rm=TRUE)
d <- median(stepsDayn,na.rm=TRUE)

msdf <-data.frame(row.names = c("without nans","with nans"),mean = c(a,c),median = c(b,d))
print(xtable(msdf), type="html")
```


The major impact of imputing the missing data is a clear increase in the number of days with few steps per day, which are often disregarded when the the nan are ignored. 

As for mean and median values, they both increase but the mean value is clearly the most affected.

# Question: Are there differences in activity patterns between weekdays and weekends?

First we add a new factor variable identifying the weekdays

```{r}
weekdays <- weekdays(as.Date(newdf$date, format="%Y-%m-%d"))

weekdays[weekdays=="sábado" | weekdays=="domingo"]<-"weekend"
weekdays[weekdays!="weekend"]<-"weekday"

newdf$weekdays <- as.factor(weekdays)

```

Now I make a panel plot containing the time series plot.

```{r}
stepsDays<-with(newdf,tapply(steps, list(interval,weekdays), sum))

par(mfrow=c(2,1))
plot(stepsDays[,1],
     type = "l",
     xlab="Intervals",
     col="red",
     main = "Steps per interval (Weekdays)",
     ylab = "Average steps"
     )
plot(stepsDays[,2],
     type = "l",
     xlab="Intervals",
     col="red",
     main = "Steps per interval (Weekend)",
     ylab = "Average steps"
     )
```















